Component response rate variation underlies the stability of complex systems (revision notes)

### 2020-01-20 notebook/revision_notes.Rmd
Figuring out variances

@@

```{r, echo = FALSE}
source("../R/random_mats.R");
source("../R/small_world.R");
rand_rho_var <- function(S, rhos, iters, int_type = 0, rmx = 0.4, C = 1, 
                         by = 1, sigma = 0.4){
    tot_res <- NULL;
    fea_res <- NULL;
    rho_res <- NULL;
    cmplxty <- NULL;
    for(i in 1:length(rhos)){
        iter           <- iters;
        tot_res[[i]]   <- matrix(data = 0, nrow = iter, ncol = 8);
        fea_res[[i]]   <- matrix(data = 0, nrow = iter, ncol = 7);
        rho_res[[i]]   <- matrix(data = 0, nrow = iter, ncol = 6);
        cmplxty[[i]]   <- matrix(data = 0, nrow = iter, ncol = 2);
        while(iter > 0){
            r_vec    <- rnorm(n = S, mean = 0, sd = rmx);
            A0       <- build_rho_mat(S = S, sigma = sigma, rho = rhos[i]);
            gam1     <- runif(n = S, min = 0, max = 2);
            A1       <- A0 * gam1;
            A0       <- A0 * mean(gam1);
            A0_stb   <- max(Re(eigen(A0)$values)) < 0;
            A1_stb   <- max(Re(eigen(A1)$values)) < 0;
            A0_fea   <- min(-1*solve(A0) %*% r_vec) > 0;
            A1_fea   <- min(-1*solve(A1) %*% r_vec) > 0;
            A0_rho   <- mat_rho(A0);
            A1_rho   <- mat_rho(A1);
            rho_diff <- A1_rho - A0_rho;
            rho_abs  <- abs(A1_rho) - abs(A0_rho);
            if(A0_stb == TRUE){
                tot_res[[i]][iter, 1] <- 1;
            }
            if(A1_stb == TRUE){
                tot_res[[i]][iter, 2] <- 1;
            }
            if(A0_fea == TRUE){
                fea_res[[i]][iter, 1] <- 1;
            }
            if(A1_fea == TRUE){
                fea_res[[i]][iter, 2] <- 1;
            }
            rho_res[[i]][iter, 1] <- A0_rho;
            rho_res[[i]][iter, 2] <- A1_rho;
            rho_res[[i]][iter, 3] <- rho_diff;
            rho_res[[i]][iter, 4] <- rho_abs;
            rho_res[[i]][iter, 5] <- max(Re(eigen(A0)$values));
            rho_res[[i]][iter, 6] <- max(Re(eigen(A1)$values));
            cmplxty[[i]][iter, 1] <- get_complexity(A0);
            cmplxty[[i]][iter, 2] <- get_complexity(A1);
            iter                  <- iter - 1;
        }
    }
    all_res     <- summarise_randmat(tot_res = tot_res, fea_res = fea_res,
                                     rho_res = rho_res, cmplxty = cmplxty);
    all_res[,1]          <- rhos;
    colnames(all_res)[1] <- "rho_set";
    return(all_res);
}

simpleboot <- function(freqs,repli=1000,alpha=0.05){    
    vals  <- NULL;                                                          
    i     <- 0;                                         
    while(i < repli){                                                   
        boot  <- sample(x=freqs,size=length(freqs),replace=TRUE);           
        strap <- mean(boot);                                                
        vals  <- c(vals,strap);                                             
        i     <- i + 1;                                                     
    }                                                                       
    vals   <- sort(x=vals,decreasing=FALSE);                            
    lowCI  <- vals[round((alpha*0.5)*repli)];                               
    highCI <- vals[round((1-(alpha*0.5))*repli)];                           
    CIs    <- c(lowCI,highCI);                                          
    return(CIs);                                                            
}

summarise_randmat <- function(tot_res, fea_res, rho_res, cmplxty){
    sims    <- length(tot_res);
    all_res <- matrix(data = 0, nrow = sims, ncol = 25);
    for(i in 1:sims){
        all_res[i, 1]  <- i + 1;
        # Stable and unstable
        all_res[i, 2]  <- sum(tot_res[[i]][,1] == FALSE);
        all_res[i, 3]  <- sum(tot_res[[i]][,1] == TRUE);
        all_res[i, 4]  <- sum(tot_res[[i]][,2] == FALSE);
        all_res[i, 5]  <- sum(tot_res[[i]][,2] == TRUE);
        # Stabilised and destabilised
        all_res[i, 6] <- sum(tot_res[[i]][,1] == FALSE & 
                                 tot_res[[i]][,2] == TRUE);
        all_res[i, 7] <- sum(tot_res[[i]][,1] == TRUE & 
                                 tot_res[[i]][,2] == FALSE);
        # Feasible and infeasible
        all_res[i, 8]  <- sum(fea_res[[i]][,1] == FALSE);
        all_res[i, 9]  <- sum(fea_res[[i]][,1] == TRUE);
        all_res[i, 10]  <- sum(fea_res[[i]][,2] == FALSE);
        all_res[i, 11]  <- sum(fea_res[[i]][,2] == TRUE);
        # Feased and defeased
        all_res[i, 12] <- sum(fea_res[[i]][,1] == FALSE & 
                                  fea_res[[i]][,2] == TRUE);
        all_res[i, 13] <- sum(fea_res[[i]][,1] == TRUE & 
                                  fea_res[[i]][,2] == FALSE);
        # Mean correlations between A[i,j] and A[j,i] off-diag elements
        all_res[i, 14] <- mean(rho_res[[i]][,1]);
        all_res[i, 15] <- mean(rho_res[[i]][,2]);
        all_res[i, 16] <- mean(rho_res[[i]][,3]);
        all_res[i, 17] <- mean(rho_res[[i]][,4]);
        all_res[i, 18] <- mean(cmplxty[[i]][,1]);
        all_res[i, 19] <- mean(cmplxty[[i]][,2]);
        all_res[i, 20] <- mean(rho_res[[i]][,5]);
        all_res[i, 21] <- mean(rho_res[[i]][,6]);
        all_res[i, 22:23] <- range(rho_res[[i]][,5]);
        all_res[i, 24:25] <- range(rho_res[[i]][,6]);
    }
    cnames <- c("N", "A0_unstable", "A0_stable", "A1_unstable", "A1_stable", 
                "A1_stabilised", "A1_destabilised", "A0_infeasible", 
                "A0_feasible", "A1_infeasible", "A1_feasible", 
                "A1_made_feasible", "A1_made_infeasible", "A0_rho", "A1_rho",
                "rho_diff", "rho_abs", "complex_A0", "complex_A1", "A0_eig",
                "A1_eig", "LCI_A0", "UCI_A0", "LCI_A1", "UCI_A1");
    colnames(all_res) <- cnames;
    return(all_res);
}
```

@@

Trying to think analytically
================================================================================

Trying to think analytically, note that the correlation $\rho$ adjusts the criteria for stability as follows [@Allesina2015a],

>>>>

Effects of correlations interactions
================================================================================

If the off-diagonal elements of $\textbf{A}$ are sampled independently from an identical distribution, then  $\max(\Re(\lambda))$ for $\textbf{M} = \textbf{A}$ can be estimated from five values [@Tang2014b]. These values include (1) system size ($S$), (2) mean self-regulation of components ($d$), (3) mean interaction strength between components ($\mu$), (4) the standard deviation between component interaction strengths ($\sigma$), and (5) the correlation of interaction strengths between components, $M_{ij}$ and $M_{ji}$ ($\rho$). When investigating the effect of varying component response rate $Var(\gamma)$ on stability by defining $\textbf{M} = \gamma\textbf{A}$, $S$ remains unchanged. Further, values of $\gamma_{i}$ were sampled such that $E[d]$ and $E[\mu]$ also remained unchanged (in practice, diagonal elements of $\textbf{M}$ were standardised so that mean values were identical before and after adding $\gamma$). What $Var(\gamma)$ does change is the variation in component interaction strengths, and $\rho$.

Variation in $\gamma$ increases the total variation such that for $\textbf{M} = \gamma \textbf{A}$, $\sigma^{2}_{M_{ij}} = \sigma^{2}_{A_{ij}} \left(1 + \sigma^{2}_{\gamma_{i}} \right)$. Eigenvalues of $\textbf{A}$ are therefore enclosed within a circle centred at $d$ with a radius of $\sigma_{A_{ij}}\sqrt{SC} = \sqrt{SC\sigma^{2}_{A_{ij}}}$, while eigenvalues of $\textbf{M}$ have a larger radius of $\sqrt{SC \sigma^{2}_{A_{ij}} \left(1 + \sigma^{2}_{\gamma_{i}}\right)}$ (see blue versus red points below).

Note that the correlation $\rho$ adjusts the criteria for stability as follows [@Allesina2015a],

@@

```{r, echo = FALSE}
source("../R/random_mats.R");
source("../R/small_world.R");
source("../R/scale_free.R");
source("../R/utilities.R");
```

@@

### 2020-01-20 notebook/revision_notes.tex
Figuring out variances

\hypertarget{role-of-correlated-matrices-in-stabilisation}{%
\section{Role of correlated matrices in
stabilisation}\label{role-of-correlated-matrices-in-stabilisation}}

In complex systems represented by large random matrices, correlation
between matrix elements \(A_{ij}\) and \(A_{ji}\) affects the
distribution of eigenvalues and therefore local stability. As the
correlation between matrix elements (\(\rho\)) decreases, the eigenvalue
spectra changes such that more variation falls along the imaginary axis.
The figure panels below compare a random matrix in which \(\rho = 0\)
(left) to one in which \(\rho\) = -0.5 (right). In both panels, complex
systems include \(S = 1000\) components, with diagonal elements of
\(-1\) and off-diagonal elements drawn from a normal distribution with a
mean of \(\mu = 0\) and standard deviation of \(\sigma = 0.4\).

\includegraphics{revision_notes_files/figure-latex/unnamed-chunk-2-1.pdf}

Because of this effect of \(\rho\) on the eigenvalue spectra, decreasing
values of \(\rho\) will also tend to decrease the rightmost eigenvalue
of the matrix \(\textbf{A}\). This makes it more likely that the complex
system represented by \(\textbf{A}\) is locally stable, as stability
occurs when all real parts of eigenvalues are negative. Note that this
elongation along the imaginary axis is also characteristic of
predator-prey communities (in which, by definition \(A_{ij}\) and
\(A_{ji}\) have opposing signs). Also note that as \(\rho\) increases
such that \(\rho > 0\), the same elongation happens along the real axis,
making random complex systems less likely to be stable.

A simple numerical analysis illustrates the linear relationship between
\(\rho\) and the expected value of the real part of the leading
eigenvalue, \(\max(\Re(\lambda))\). Below, I have run 10000 simulations
across values of \(\rho\) from -0.9 to 0.9 for complex systems with
\(S = 25\) components. Error bars show 95\% bootstrapped confidence
intervals.

\includegraphics{revision_notes_files/figure-latex/unnamed-chunk-4-1.pdf}

In the main text, I demonstrated that when \(S\) is finite but system
complexity \(\sigma\sqrt{SC}\) is high (\(C\) defines the connectance of
\(\textbf{A}\), or the proportion of non-zero off-diagonal elements),
variation in component response rate \(\gamma\) often underlies system
stability. In other words, highly complex systems that are observed to
be stable would typically not be if we removed the variation in their
component response rates. Mathematically, this means that by multiplying
\(\textbf{A}\) by a diagonal matrix \(\gamma\) with variable elements,
the sign of \(\max(\Re(\lambda))\) is sometimes flipped from positive to
negative in these finite systems of high complexity. Interestingly, this
increase in stability given \(Var(\gamma) > 0\) is not necessarily
caused by \(\gamma\) decreasing \(\rho\). In fact, for \(S = 25\),
\(\sigma = 0.4\), and \(C = 1\), random complex systems that are
stabilised by \(\gamma\) typically have increased \(\rho\) values. Note
that for these parameter values, 1 million simulations found
\(\textbf{M} = \gamma \textbf{A}\) to be stable for 36 systems when all
\(\gamma_{i} = 1\), but 383 systems when
\(\gamma_{i} \sim \mathcal{U}(0, 2)\). Below shows the distribution of
the difference in \(\rho\) between systems with versus without
\(var(\gamma)\) for 1000000 stabilised systems; that is,
\(\rho_{change} = \rho_{var(\gamma)} - \rho_{\gamma = 1}\).

\includegraphics{revision_notes_files/figure-latex/unnamed-chunk-7-1.pdf}

When \(\textbf{A}\) was stabilised by \(Var(\gamma)\), the change in
\(\rho\) was normally distributed around
\ensuremath{4.8394472\times 10^{-6}} (95\% CIs:
\ensuremath{-7.3124139\times 10^{-5}},
\ensuremath{8.7487968\times 10^{-5}}). Hence, decreasing the correlation
between \(\textbf{A}_{ij}\) and \(\textbf{A}_{ji}\) was not by itself
the cause of stability. For a clearer picture of the effect of
\(\gamma\), it is useful to show the relationship between \(\rho\) and
\(\max(\Re(\lambda))\) again as above, but this time also for how the
relationship changes given \(Var(\gamma)\). That is, given
\(\textbf{M} = \textbf{A}\) in addition to
\(\textbf{M} = \gamma\textbf{A}\).

\includegraphics{revision_notes_files/figure-latex/unnamed-chunk-8-1.pdf}

Including \(Var(\gamma)\) introduces a nonlinear relationship between
\(\rho\) and \(\max(\Re(\lambda))\). Points along the x-axis are spaced
more closely together given \(Var(\gamma)\) because \(Var(\gamma)\)
tends to decrease the absolute magnitude of \(\rho\). Grey and red
points centred on \(\rho = 0\) represent the same simulations, before
(grey) and after (red) including \(Var(\gamma)\). Grey and red points to
the left and right show decreasing and increasing simulated \(\rho\)
values, respectively. Faint grey lines connect points for the same set
of simulations, and where the red point is lower than the black point,
the expected \(\max(\Re(\lambda))\) was lower given \(Var(\gamma)\).

The region of \(\rho\) values for \(\textbf{A}\) that would result in
increased stability given \(Var(\gamma)\) is highlighted with red
shading below.

\includegraphics{revision_notes_files/figure-latex/unnamed-chunk-9-1.pdf}

In this red shaded region above, \(\max(\Re(\lambda))\) is decreased by
\(Var(\gamma)\). The shaded region encompasses values of \(\rho\)
between -0.4 and 0.7. But 1000000 simulated \(\textbf{A}\) had values
that ranged between -0.2642846 and 0.2668648, meaning that
\(\max(\Re(\lambda))\) was always expected to decrease given
\(Var(\gamma)\) even if \(Var(\gamma)\) caused \(\rho\) to increase.

The curvature of the relationship between \(\rho\) and
\(\max(\Re(\lambda))\) is consistent across different values of \(S\),
as shown below. Including \(\gamma\) always results in a concave upward
relationship between \(\rho\) and \(\max(\Re(\lambda))\).

\includegraphics{revision_notes_files/figure-latex/unnamed-chunk-10-1.pdf}

In all panels above, the solid line shows the relationship between the
expected \(\rho\) between \(\textbf{A}\) elements \(A_{ij}\) and
\(A_{ji}\) given no variation in component response rates (i.e., the
diagonal matrix equals the identity matrix, \(\gamma = \textbf{I}\), so
\(\textbf{M} = \textbf{A}\)). The dotted line shows the same
relationship given variation in component response rates (i.e., the
diagonal matrix contains elements drawn from a random uniform
distribution between 0 and 2, so \(\textbf{M} = \gamma\textbf{A}\)).

\hypertarget{trying-to-think-analytically}{%
\section{Trying to think
analytically}\label{trying-to-think-analytically}}

Trying to think analytically, note that the correlation \(\rho\) adjusts
the criteria for stability as
follows\textsuperscript{\protect\hyperlink{ref-Allesina2015a}{1}},

\[\sigma\sqrt{SC}\left(1 + \rho\right) < 1.\]

Given that diagonal values of \(\textbf{M}\) are set to \(-1\), We can
know the relationship \(S = 1600\), \(C = 1\), and \(\sigma = 1/40\).
Under these parameter values, given \(\rho = 0\), we know that
\(\sigma\sqrt{SC}\left(1 + \rho\right) = 1\). Hence, the expected value
of \(\max(\Re(\lambda)) = 0\)

\begin{verbatim}
## Warning in xy.coords(x, y, xlabel, ylabel, log): imaginary parts discarded
## in coercion
\end{verbatim}

\includegraphics{revision_notes_files/figure-latex/unnamed-chunk-11-1.pdf}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Allesina2015a}{}%
1. Allesina, S. \& Tang, S. The stability--complexity relationship at
age 40: a random matrix perspective. \emph{Population Ecology} 63--75
(2015).
doi:\href{https://doi.org/10.1007/s10144-014-0471-0}{10.1007/s10144-014-0471-0}

>>>>

\hypertarget{effects-of-correlations-interactions}{%
\section{Effects of correlations
interactions}\label{effects-of-correlations-interactions}}

If the off-diagonal elements of \(\textbf{A}\) are sampled independently
from an identical distribution, then \(\max(\Re(\lambda))\) for
\(\textbf{M} = \textbf{A}\) can be estimated from five
values\textsuperscript{\protect\hyperlink{ref-Tang2014b}{1}}. These
values include (1) system size (\(S\)), (2) mean self-regulation of
components (\(d\)), (3) mean interaction strength between components
(\(\mu\)), (4) the standard deviation between component interaction
strengths (\(\sigma\)), and (5) the correlation of interaction strengths
between components, \(M_{ij}\) and \(M_{ji}\) (\(\rho\)). When
investigating the effect of varying component response rate
\(Var(\gamma)\) on stability by defining
\(\textbf{M} = \gamma\textbf{A}\), \(S\) remains unchanged. Further,
values of \(\gamma_{i}\) were sampled such that \(E[d]\) and \(E[\mu]\)
also remained unchanged (in practice, diagonal elements of
\(\textbf{M}\) were standardised so that mean values were identical
before and after adding \(\gamma\)). What \(Var(\gamma)\) does change is
the variation in component interaction strengths, and \(\rho\).

Variation in \(\gamma\) increases the total variation such that for
\(\textbf{M} = \gamma \textbf{A}\),
\(\sigma^{2}_{M_{ij}} = \sigma^{2}_{A_{ij}} \left(1 + \sigma^{2}_{\gamma_{i}} \right)\).
Eigenvalues of \(\textbf{S}\) are therefore enclosed within a circle
centred at \(d\) with a radius of
\(\sigma_{A_{ij}}\sqrt{SC} = \sqrt{SC\sigma^{2}_{A_{ij}}}\), while
eigenvalues of \(\textbf{A}\) have a larger radius of
\(\sqrt{SC\sigma^{2}_{A_{ij}} \left(1 + \sigma^{2}_{\gamma_{i}\right)}}\)

Note that the correlation \(\rho\) adjusts the criteria for stability as
follows\textsuperscript{\protect\hyperlink{ref-Allesina2015a}{2}},

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-Tang2014b}{}%
1. Tang, S. \& Allesina, S. Reactivity and stability of large
ecosystems. \emph{Frontiers in Ecology and Evolution} \textbf{2,} 1--8
(2014).

\leavevmode\hypertarget{ref-Allesina2015a}{}%
2. Allesina, S. \& Tang, S. The stability--complexity relationship at
age 40: a random matrix perspective. \emph{Population Ecology} 63--75
(2015).
doi:\href{https://doi.org/10.1007/s10144-014-0471-0}{10.1007/s10144-014-0471-0}
